---
title: "UASE"
output:
  html_document: default
  pdf_notebook: default
  pdf_document: default
---

```{r setup, include=FALSE}
library(ggplot2)
knitr::opts_chunk$set(message = FALSE)
theme_set(theme_bw())

```


## Without Mixtures

We first load the ClaritySim and Clarity package and simulate some appropriate data:

```{r}
library(ClaritySim)
library(Clarity)
set.seed(1)
n=100 
k=10
original <- simulateCoalescent(n,k,
                         sigma0=0.0001,
                         Amodel="uniform",
                         alpha=0,
                         minedge=0.1) # Simlulate 100 objects in a 10 dimensional latent space

similar <- mixCoalescent(original,fraction=1)


     
labels <- original$A %*% as.vector(1:10)
```

We can now visualize the trees from these 3 different simulated data:

```{r}
par(mfrow = c(1,2))
plot(original$tree)
plot(similar$tree)
```

We now obtain the similarity matrices for the subjects in both models and get a joint matrix 

```{r}
original_dist <- original$Y

similar_dist <- similar$Y

joint_dist <- cbind(original_dist, similar_dist)
```

## UASE 

We want to perform obtain the UASE right and left embeddings of the joint similarity matrix. First write the corresponding functions:
```{r}
UASE_right <- function(A,d = 2, K){
  svd <- svd(A)
  Y <- svd$v %*% diag(sqrt(svd$d))
  Y <- Y[,1:d]
  t <- rep(1:K, each = nrow(A))
  data <- rep(1:nrow(A),K)
  
  Y <- as.data.frame(cbind(Y,as.factor(t), as.factor(data)))
  colnames(Y) <- c(1:d,"discipline", "data_point")
  return(Y)
}

UASE_left <- function(A,d = 2, K){
  svd <- svd(A)
  Y <- svd$u %*% diag(sqrt(svd$d))
  Y <- Y[,1:d]
  Y <- as.data.frame(Y)
  colnames(Y) <- c(1:d)
  return(Y)
}
```

We now get the embeddings
```{r}
library(dplyr)
library(ggplot2)
library(viridis)
library(gridExtra)

U <- UASE_left(joint_dist, K = 2)
V <- UASE_right(joint_dist, K= 2)

U <- U %>%
  as_tibble() %>%
  mutate(labels = as.factor(labels))

original_V <- V[1:n,] %>%
  as_tibble() %>%
  mutate(labels = as.factor(labels))


similar_V <- V[-(1:n),] %>%
  as_tibble() %>%
  mutate(labels = as.factor(labels))

original_V_plot <- ggplot(original_V, aes(`1`,`2`, color = labels)) +
  geom_point(alpha = 0.5) +
  geom_jitter() +
  scale_color_viridis_d(option = "turbo") + 
  theme(legend.position = "none") +
  labs(title = "V Discipline 1")

similar_V_plot <- ggplot(similar_V, aes(`1`,`2`, color = labels)) +
  geom_point(alpha = 0.5) +
  geom_jitter() +
  scale_color_viridis_d(option = "turbo") + 
  theme(legend.position = "none") +
  labs(title = "V Discipline 2")

U_plot <- ggplot(U, aes(`1`,`2`, color = labels)) +
  geom_point(alpha = 0.5) +
  geom_jitter() +
  scale_color_viridis_d(option = "turbo", guide = guide_legend(ncol = 3, label.theme = element_text(size= 8))) +
  labs(title = "U")

lay <- rbind(c(1,2),
             c(3,3))
grid.arrange(original_V_plot, similar_V_plot, U_plot, layout_matrix = lay)
```

   
```{r eval=FALSE, include=FALSE}
U_groups <- U %>% 
  filter(labels == c(1,2,3,4))

ggplot(U_groups, aes(`1`,`2`, color = labels)) +
  geom_point(alpha = 0.2) +
  geom_jitter()  +
  scale_color_viridis_d(option = "turbo", guide = guide_legend(ncol = 3, label.theme = element_text(size= 8))) 
  
```

### Clustering 
Get UPMGA tree from U with different values of d

```{r}
library(ggdendro)
library(dendextend)

D <- 5
par(mfcol = c(2,3))
plot(original$tree)
plot(similar$tree)
UPGMA_plots <- vector(mode = "list", D-1)
for (i in 2:D){
  U_d <- UASE_left(joint_dist, d = i,K = 2)
  U_dist <- dist(U_d[,1:i])
  
  U_UPGMA <- hclust(U_dist, method = "average", members = labels)
  dend <- as.dendrogram(U_UPGMA)
  labels_colors(dend) <- labels[U_UPGMA$order]
  plot(dend, horiz = TRUE)
  title(paste("d =", i))
}



```





## With Mixtures

```{r}
mixture <- mixCoalescent(original)

mixture_dist <- mixture$Y
```
