---
title: "Linux Command Line"
author: "Rachel Wood"
date: "2023-03-22"
output: 
  pdf_document: 
    highlight: espresso
header-includes: 
  - \usepackage{tikz}
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
library(formatR)
library(knitr)
library(highr)

hook_output = knit_hooks$get('source',)
opts_knit$set(dev = 'pdf',
              pdf.options(encoding = "ISOLatin9.enc"),
              message = FALSE,
              warning = FALSE,
              tidy.opts = list(width.cutoff=50), tidy=TRUE,
              fig.pos ="H", out.width = '60%',
              fig.align = 'center',
              root.dir = "/home/ac18826/Documents/COMPASS/First_Year/TB2/Statistical_Computing/Intro to Linux/Portfolio/"
              )
```

For this portfolio I will use the example of bioinformatics to showcase the utility of the Linux command line. The data we will be using is the genome and protein datasets from the SARS-CoV-2 virus, which can be found at \url{https://www.ncbi.nlm.nih.gov/genome/?term=SARS-CoV-2}.

# Working with Files and File Systems

```{r, engine='bash', include = FALSE}
rm -r bioinf_ex
cd ..
cp GCF_009858895.2_ASM985889v3_genomic.fna.gz Portfolio/
cp GCF_009858895.2_ASM985889v3_protein.faa.gz Portfolio/

```


We start in a folder containing the files downloaded from the link above:

```{r, engine='bash', eval = FALSE}
ls
GCF_009858895.2_ASM985889v3_genomic.fna.gz  
GCF_009858895.2_ASM985889v3_protein.faa.gz
```

We first create a folder to work within using the `mkdir` command and use the `cd` command to move into this new file. Finally we can use the `ls` command to view the contents of this folder:
```{r, engine='bash'}
mkdir bioinf_ex

```

As expected, the directory is empty, but we can move the genome and protein files to our new folders. We use the \texttt{cd ..} to go one `step back' in the directory and return to our original directory. We use the \texttt{ls} command again and then the \texttt{mv} command to move our two files into the new folder we have created:




```{r, engine='bash'}
mv GCF_009858895.2_ASM985889v3_protein.faa.gz -t bioinf_ex/
mv GCF_009858895.2_ASM985889v3_genomic.fna.gz -t bioinf_ex/
cd bioinf_ex/
```
```{r, include=FALSE}
opts_knit$set(root.dir ="/home/ac18826/Documents/COMPASS/First_Year/TB2/Statistical_Computing/Intro to Linux/Portfolio/bioinf_ex/")
```
We now check these files have been successfully moved:
```{r, engine ='bash', eval = FALSE}
ls 
GCF_009858895.2_ASM985889v3_genomic.fna.gz  
GCF_009858895.2_ASM985889v3_protein.faa.gz
```


We notice the `.gz` file extension, meaning we need to unzip the files. We can do this with the `gzip` command:

```{r, engine = 'bash'}
gzip -d GCF_009858895.2_ASM985889v3_genomic.fna.gz
gzip -d GCF_009858895.2_ASM985889v3_protein.faa.gz
```

```{r, engine='bash', eval = FALSE}
ls
GCF_009858895.2_ASM985889v3_genomic.fna
GCF_009858895.2_ASM985889v3_protein.fa
```


