---
title: "LA with RcppArmadillo"
author: "Rachel Wood"
date: "2023-03-28"
output:
  html_document:
    df_print: paged
  pdf_document: 
    highlight: tango
header-includes: 
  - \usepackage{tikz}
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}} 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(ggplot2)
library(viridis)
theme_set(theme_bw())
scale_fill_continuous <- function(){
  scale_fill_viridis(option = "plasma", begin = 0.2)
}
knitr::opts_knit$set(message = FALSE,
              warning = FALSE,
              tidy.opts = list(width.cutoff=50), tidy=TRUE,
              fig.pos ="H", out.width = '60%',
              fig.align = 'center'
              )
```

This portfolio aims to perform smoothing with RcppArmadillo, using local polynomial regression. 

We load the relevant data on solar power production in Australia:
```{r}
library(dplyr)
load("solarAU.RData")
head(solarAU)
solarAU <- as_tibble(solarAU)
```
We work with the log of the production as it is less skewed (adding 0.01 to all entries to avoid numerical errors)
```{r}
solarAU <- solarAU %>%
  mutate(logProduction = log(prod + 0.01))
```


We can now plot the production in a 2d summary:
```{r}
library(ggplot2)
library(viridis)
ggplot(solarAU,
       aes(x = toy, y = tod, z = logProduction)) +
       stat_summary_2d() 
```

We can see there is greater solar production in the middle of the day and in the winter, which is to be expected. 

# Linear Model

We consider the following model:
$$
\mathbb{E}(y|x) = \beta_0 + \beta_1\text{tod} + \beta_2\text{tod}^2 + \beta_3\text{toy} + \beta_4\text{toy}^2 
$$
We can use R to solve this easily:
```{r}
fit <- lm(logProduction ~ tod + I(tod^2) + toy + I(toy^2), data = solarAU)
```

## Using RcppArmadillo
We now want to use RcppArmadillo to fit the same model, taking as input 
```{r}
X <- with(solarAU)
```




