---
title: "LA with RcppArmadillo"
author: "Rachel Wood"
date: "2023-03-28"
output:
  html_document:
    df_print: paged
    highlight: breezedark
  pdf_document: 
    highlight: tango
header-includes: 
  - \usepackage{tikz}
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}} 
  - \DeclareMathOperator*{\argmin}{arg\,min}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(ggplot2)
library(viridis)
theme_set(theme_bw())
scale_fill_continuous <- function(){
  scale_fill_viridis(option = "plasma", begin = 0.2)
}
knitr::opts_knit$set(message = FALSE,
              warning = FALSE,
              tidy.opts = list(width.cutoff=50), tidy=TRUE,
              fig.pos ="H", out.width = '60%',
              fig.align = 'center'
              )
```

This portfolio aims to perform smoothing with RcppArmadillo, using local polynomial regression. 

We load the relevant data on solar power production in Australia:
```{r}
library(dplyr)
load("solarAU.RData")
head(solarAU)
solarAU <- as_tibble(solarAU)
```
We work with the log of the production as it is less skewed (adding 0.01 to all entries to avoid numerical errors)
```{r}
solarAU <- solarAU %>%
  mutate(logProduction = log(prod + 0.01))
```


We can now plot the production in a 2d summary:
```{r}
library(ggplot2)
library(viridis)
ggplot(solarAU,
       aes(x = toy, y = tod, z = logProduction)) +
       stat_summary_2d() 
```
We can see there is greater solar production in the middle of the day and in the winter, which is to be expected. 

# Linear Model

We consider the following model:
$$
\mathbb{E}(y|x) = \beta_0 + \beta_1\text{tod} + \beta_2\text{tod}^2 + \beta_3\text{toy} + \beta_4\text{toy}^2 
$$
We can use R to solve this easily:
```{r}
fit <- lm(logProduction ~ tod + I(tod^2) + toy + I(toy^2), data = solarAU)
R_beta <- fit$coefficients
```

## Using RcppArmadillo
We now want to use RcppArmadillo to fit the same model, taking as input 
```{r}
X <- with(solarAU, cbind(1, tod, tod^2, toy, toy^2))
y <- solarAU$logProduction
```


### QR decomposition
We first try to perform the computation using a QR decomposition
```{r, engine='Rcpp'}
// [[Rcpp::depends(RcppArmadillo)]]

#include <RcppArmadillo.h>
using namespace arma;

// [[Rcpp::export]]
vec arma_QR_lm(mat& X, vec& y){
  mat Q;
  mat R;
  
  qr(Q, R, X);
  vec Qty = trans(Q) * y;
  
  vec beta = solve(R, Qty);
  return beta;
}
```



We now check that the two vectors are the same:
```{r}
arma_QR_beta <- arma_QR_lm(X,y)
max(abs(R_beta - arma_QR_beta))
```

We can now use the microbenchmark function to compare the running times
```{r, cache=TRUE}
library(microbenchmark)
microbenchmark(R = beta <- lm(logProduction ~ tod + I(tod^2) + toy + I(toy^2), data = solarAU),
               "arma QR"= arma_QR_beta <- arma_QR_lm(X,y), times = 10)
```

We can see here however that the `arma` function performed much worse than the base `R` one. This might be because we have used `qr` instead of `qr_econ` in our Rcpp code. We can rewrite the Rcpp function
```{r, engine='Rcpp'}
// [[Rcpp::depends(RcppArmadillo)]]

#include <RcppArmadillo.h>
using namespace arma;

// [[Rcpp::export]]
vec arma_QR_lm_fast(mat& X, vec& y){
  mat Q;
  mat R;
  
  qr_econ(Q, R, X);
  vec Qty = trans(Q) * y;
  
  vec beta = solve(R, Qty);
  return beta;
}
```

```{r, cache=TRUE}
library(microbenchmark)
microbenchmark(R = R_beta <- lm(logProduction ~ tod + I(tod^2) + toy + I(toy^2), data = solarAU)$coefficients,
               "arma QR"= arma_QR_beta <- arma_QR_lm(X,y), 
               "arma QR fast" = arma_QR_beta_fast <- arma_QR_lm_fast(X,y),
               times = 10)
```
We can now see, the RcppArmadillo version outperforms the R version, however we want to see if this comes at a cost of precision:
```{r}
arma_QR_beta_fast <- arma_QR_lm_fast(X,y)
max(abs(arma_QR_beta_fast - R_beta))
```

### Cholesky Decomposition

**REVISIT??**

```{r, engine='Rcpp', eval = FALSE}
// [[Rcpp::depends(RcppArmadillo)]]

#include <RcppArmadillo.h>
using namespace arma;

// [[Rcpp::export]]
vec arma_cholesky_lm(mat& X, vec& y){
  
}
```

## Model Checking
We can now plot fitted values and the residuals to check if the model seems valid:

```{r}
library(gridExtra)

solarAU <- solarAU %>%
  mutate(fitted = fit$fitted.values, 
         residuals = fit$residuals) 

fitted_plot <- ggplot(data = solarAU,
                      aes(x = toy, y = tod, z = fitted)) +
  stat_summary_2d()+
  theme(aspect.ratio = 1)

res_plot <- ggplot(data = solarAU,
                      aes(x = toy, y = tod, z = residuals)) +
  stat_summary_2d()+
  theme(aspect.ratio = 1)

grid.arrange(fitted_plot, res_plot, ncol = 2)
```
We can see from these plots there is a non linear pattern, particularly in the residuals, so we adapt the model.

# Kernel Local Least Regression

In this model we take the coefficients to be dependent on the covariates $\hat \beta = \hat \beta(x)$. For a fixed $x_0$, we take
$$
\hat \beta (x_0) = \argmin_{\beta} \sum_{i =1}^n \kappa_H (x_0 - x_i) (y_i - \tilde x_i^T \beta)^2
$$

for a kernel $\kappa_H$ with bandwidth H. We create this function in R:

```{r}
library(mvtnorm)

lm_local <- function(y, x0, X0, x, X, H){
  w <- dmvnorm(x, x0, H)
  fit <- lm(y ~ -1 + X, weights = w)
  return( t(X0) %*% coef(fit) )
}
```

To reduce the computational cost, we will only fit the model on 2000 data points:

```{r}
sample_ind <- sample(1:nrow(solarAU), 2000)

x <- solarAU %>%
  select(tod, toy) %>%
  as.matrix()
x_sample <- x[sample_ind,] 
X_sample <- X[sample_ind,]

y_sample <- solarAU$logProduction

solar_sample <- solarAU[sample_ind,]

X_sample <- X[sample_ind,]
```

We can now fit $\hat \beta (x_0)$ for each sample data point:
```{r}
pred_local <- sapply(1:2000, function(ii){
  lm_local(y = y, x0 = x_sample[ii, ], X0 = X_sample[ii, ], x = x, X = X, H = diag(c(1, 0.1)^2))
})
```

```{r}
solar_sample <- solar_sample %>%
  mutate("local fitted" = pred_local,
         residuals = logProduction - `local fitted`)

local_fitted_plot <- ggplot(solar_sample,
       aes(x = toy, y = tod, z = `local fitted`)) +
       stat_summary_2d() +
       theme(aspect.ratio = 1)

local_res_plot <- ggplot(solar_sample,
       aes(x = toy, y = tod, z = residuals)) +
       stat_summary_2d() +
       theme(aspect.ratio = 1)

grid.arrange(local_fitted_plot, local_res_plot, ncol = 2)
```

## Using RcppArmadillo

We now want to implement the above with RcppArmadillo in the hopes of speeding this up 

```{r, engine='Rcpp'}
// [[Rcpp::depends(RcppArmadillo, RcppDist)]]

#include <RcppArmadillo.h>
#include <mvnorm.h>
using namespace arma;

vec arma_weight_lm(mat& X, vec& y, vec& w){
  mat W = diagmat(sqrt(w));
  
  mat new_X = W * X;
  vec new_y = W * y;
  
  mat Q;
  mat R;
  
  qr_econ(Q, R, new_X);
  vec Qty = trans(Q) * new_y;
  
  vec beta = solve(R, Qty);
  return beta;
}

// [[Rcpp::export]]
vec arma_lm_local(vec& y, vec& x0,  vec& X0, mat& x, mat& X, mat& H){

  vec w = dmvnorm(x, x0, H);
  
  vec beta = arma_weight_lm(X,  y, w);
  
  return trans(X0) * beta;
  
  return y;
}

```

```{r}
arma_pred_local <- sapply(1:2000, function(ii){
  arma_lm_local(y = y, x0 = x_sample[ii, ], X0 = X_sample[ii, ], x = x, X = X, H = diag(c(1, 0.1)^2))
})
```

