---
title: "Common R Functions"
author: "Rachel"
date: "2022-10-23"
output: pdf_document
---
This portfolio will detail the different ways we can make code faster and more efficient by eliminating loops as much as possible. We will first define a few functions containing loops, for which we will write more efficient versions. We can then compare the speed of these using the `system.time()` function.

We start with some basic functions with vectorised base R analogues (we do this for emphasis, as the results for this are more dramatic than for a vector):
```{r}
sum_loop <- function(x){
  count <- 0
  for (i in 1:nrow(x)){
    for (j in 1:ncol(x)){
      count <- count + x[i,j]
    }
  }
  return(count)
}
```

```{r}
sin_loop <- function(x){
  sin_mat <- matrix(nrow = nrow(x), ncol = ncol(x))
  for (i in 1:nrow(x)){
    for (j in 1:ncol(x)){
      sin_mat[i,j] <- sin(x[i,j])
    }
  }
  return(sin_mat)
}
```

We can then consider some more complicated functions. 

# Vectorisation
Many base R functions are vectorised, meaning they will take a vector as input and will perform the computation on the vector elementwise, which will run much faster than if we were to use a loop. To see this, we can compare the run times for our `sum_loop()` function and the built-in `sum()` function on an arbitrary matrix `X`:
```{r}
x <- matrix(1:1000000, nrow =1000)
system.time(sum_loop(x))
system.time(sum(x))
```
While there is a difference, it might be more obvious if we apply two nested functions, each containing a loop and two nested vectorised functions:
```{r}
system.time(sum_loop(sin_loop(x)))
system.time(sum(sin(x)))
```


# Apply Family of Functions
This class of built-in R functions allows to apply the same function across lists, arrays or dataframes.

# Parallel Programming





create final plot of length of vector vs time??