---
title: "SM2 Assessed Coursework"
author: "Rachel Wood"
date: "2023-03-14"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

```{=tex}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\argmax}[1]{\underset{#1}{\operatorname{argmax}}}
```
This portfolio considers a model for $n$ observations $\{ (y_i^0, x_i^0)\} \in \mathbb{R} \times \mathbb{R}^{p}$ defined by

\begin{equation} \label{model}
Y_i^0 \sim f(y;\mu_i, \phi)dy, \hspace{0.5cm} g(\mu_i) = \alpha + f(x_i^0), \hspace{0.5cm} \text{for} i = 1, \ldots, n
\end{equation} where $\alpha \in \mathbb{R}, \ \phi \in (0, \infty)$ and $f \in \mathcal{F} = \mathcal{H}_k$.

Here $(\mathcal{H_k}, \left< \cdot, \cdot \right>)$ is a reproducible kernel Hilbert space (RKHS) with positive semi definite kernel $k$. Then we have that $\mathcal{H}_k$ satisfies the reproducibility property

```{=tex}
\begin{equation}
f(x) = \left< f, k(x, \cdot ) \right> \hspace{0.5cm} \forall x \in \mathcal{X}, \forall f \in \mathcal{H}_k
\end{equation}
```
# Part 1

## Question 1

Here we consider the identifiability of (\ref{model}). A kernel will lead to identifiable models if and only if the corresponding RKHS does not contain constant functions.

\paragraph{Unidentifiable kernel:}

We can take the kernel to be $$
k(x,y) = 1
$$ which is positive semi-definite, we can see that taking $\H_k = \{ f: f(x) = c \ \forall x \}$ and $\left< f , g\right>_k = fg$ satisfies the reproducing property: $$
f(x) = \left< f,\  k(x, \cdot) \right> = \left< f, 1 \right> = f \hspace{0.5cm} \forall x \in \mathcal{X}, \ \forall f \in \mathcal{H}_k
$$ and thus $(\mathcal{H}_k, \left< \cdot, \cdot \right>_k)$ is the corresponding RKHS. Since $\mathcal{H}_k$ is made up of constant functions, $k$ does not produce identifiable models.

\paragraph{Identifiable Kernel:}

An example of an identifiable kernel is the Gaussian kernel: $$
k_\lambda ( x, x')= \exp \left( - \frac{||x - x' ||^2}{\lambda} \right) 
$$ Any function in the corresponding RKHS $f \in \mathcal{H}_k$ must satisfy: $$
f(x) = \left< f, k(x, \cdot)\right> = \frac{1}{\lambda} \left< f, - \exp||x - \cdot  ||^2 \right>
$$ It is clear that there is no constant function $f$ (excepting the zero function) which satisfies this. Hence the Gaussian kernel leads to identifiable models.

## Question 2

For this question, we consider the solution $\hat f_\lambda$ to the optimisation problem

```{=tex}
\begin{equation} \label{opt}
  ( \hat{\alpha_\lambda}, \hat{\phi_\lambda}, \hat{f_\lambda} ) \in \argmax{\alpha \in \mathbb{R}, \phi \in (0, \infty), f \in \H_k} \frac{1}{2n} \ \sum_{i = 1}^{n} \log f \left( y_i ; g^{-1} \left( \alpha + f(x_i^0) \right), \phi \right) - \lambda ||f||_{\H_k}^2
\end{equation}
```
We assume $\mathcal{H}_k = \tilde{\mathcal{H}_{n}} \oplus \tilde{\mathcal{H}_{n}}^{\perp}$, hence we can write $\hat f_\lambda = f_1 + f_2$ where $f_1 \in \tilde{\mathcal{H}_n}$ and $f_2 \in \tilde{\mathcal{H}_n}^{\perp}$. Thus there exists coefficients $\hat \beta_\lambda = \left( \hat \beta_{\lambda,1}, \ldots, \hat \beta_{\lambda,n} \right) \in \mathbb{R}^n$ such that

$$
f_1 = \sum_{i = 1}^n \hat \beta_{\lambda,i} k(x_i^0, \cdot)
$$

and we can write $f = \sum_{i = 1}^n \hat \beta_{\lambda,i} k(x_i^0, \cdot) + f_2$. Further, by the reproducing property, for any $x_j$:

$$
f(x_j) = \left< \sum_{i = 1}^n \hat \beta_{\lambda,i} k(x_i^0, \cdot) + f_2, \ k(x_j^0,\cdot) \right>  = \sum_{i = 1}^n \hat \beta_{\lambda,i} \left< k(x_i^0, \cdot), k(x_j^0, \cdot) \right> = \sum_{i = 1}^n \hat \beta_{\lambda,i} k(x_i^0, x_j^0)
$$

and so $f(x_j)$ does not depend on $f_2$ and as a consequence, the first term in (\ref{opt}) also does not depend on $f_2$. Hence to choose $f_2$ we only need to consider minimizing the regularisation term. Using that $\left<f_1, f_2 \right> = 0$, we see

$$
||f||_{\mathcal{H}_k}^2 = \left< f_1 +f_2, f_1 +f_2 \right> = ||f_1||_{\mathcal{H}_k}^2 + ||f_2||_{\mathcal{H}_k}^2 \geq ||f_1||_{\mathcal{H}_k}^2 
$$

with the last inequality becoming an equality when $f_2 = 0$. Hence the minimiser $\hat f_\lambda$ must have $f_2 = 0$ and can be written as

```{=tex}
\begin{equation} \label{f_sol}
  \hat f_\lambda = \sum_{i = 1}^n \hat \beta_{\lambda,i} k(x_i^0, \cdot)
\end{equation}
```
## Question 3

We now want to substitute the results of (\ref{f_sol}) into (\ref{opt}). The regularisation term becomes:

```{=tex}
\begin{align*}
||f||_{\H_k}^2 &=  \left< \sum_{i = 1}^n \hat \beta_{\lambda,i} k(x_i^0, \cdot), \sum_{i = j}^n \hat \beta_{\lambda,j} k(x_j^0, \cdot) \right> \\
&= \sum_{i = 1}^{n} \sum_{j = 1}^n \hat \beta_{\lambda,i} \left< k(x_i^0, \cdot), k(x_j^0, \cdot) \right> \hat \beta_{\lambda,j} \\
&= \sum_{i = 1}^{n} \sum_{j = 1}^n \hat \beta_{\lambda,i} k(x_i^0, x_j^0) \hat \beta_{\lambda,j} \\
&= \hat \beta_\lambda^T K \hat \beta_\lambda
\end{align*}
```
and so (\ref{opt}) becomes

```{=tex}
\begin{equation} \label{new_opt}
(\hat{\alpha_\lambda}, \hat{\phi_\lambda}, \hat{\beta_\lambda} ) \in \argmax{\alpha \in \mathbb{R}, \phi \in (0, \infty), \beta \in \mathbb{R}^n} \ \frac{1}{2n} \sum_{i = 1}^{n} \log f \left( y_i ; g^{-1} \left( \alpha + \sum_{j = 1}^n \beta_{i} k(x_i^0, x_j^0) \right), \phi \right) - \lambda \beta^T K \beta
\end{equation}
```
## Question 4

Given $m \leq n+2$, we want to obtain an $m$-dimensional problem. Then $d = m -2$ is the length of the new $\tilde \beta_\lambda$ vector to estimate.

The Nyström method approximates $k$ by $\tilde{f}^{(m)} = k^{(m)}$, given by 

$$
\tilde k^{(d)} ( x, x') = k_d (x)^T  K_{d,d}^{-1} \ k_d(x')
$$ 

where $K_{d, d} \in \mathbb{R}^{d\times d}$ is the first $d$ rows and columns of the gram matrix $K$ and $k_d (x) = (k(x_1^0,x), \ldots, k(x_d^0, x))$

Then we can write $f(x)$ as $$
f =\sum_{i = 1}^n \beta_{i}\ \tilde k_d(x_i^0, \cdot) = \sum_{i = 1}^n \beta_{i}\ k_d (x_i^0)^T K_{d,d}^{-1} \ k_d(\cdot) = \left( \sum_{i = 1}^n \beta_{i}\ k_d (x_i^0)^T  K_{d,d}^{-1} \right) k_d(\cdot)
$$

and take the new vector of coefficients to be: $$
\tilde \beta^T = \sum_{i = 1}^n \beta_{i}\ k_d (x_i^0)^T \left( K_d^0 \right)^{-1} = \beta^T K_{n,d} K_{d,d}^{-1}
$$ 

where $K_{n,d} \in \mathbb{R}^{n \times d}$ is the first $d$ columns of $K$ We can now rewrite $f$ as: 

$$
f = \tilde \beta^T k_d(\cdot) = \sum_{j =1}^d \tilde \beta_j  k(x_j, \cdot)
$$

We now consider the regularisation term, again we substitute $\tilde k^{(d)}$:

$$
\beta^T \tilde K^{(d)} \beta = \beta^T K_{n,d}\  K_{d,d}^{-1} \ K_{n,d}^T \ \beta = \left(\beta^T K_{n,d} \ K_{d,d}^{-1}   \right) K_{d,d}^T \left( K_{d,d}^{-T}\ K_{n,d}\ \beta \right) = \tilde \beta^T K_{d,d}^T \ \tilde \beta
$$ 

Hence we can now write our $m$-dimension optimisation problem:

```{=tex}
\begin{equation} \label{m_opt}
(\hat{\alpha_\lambda}, \hat{\phi_\lambda}, {\tilde \beta_\lambda} ) \in \argmax{\alpha \in \mathbb{R}, \phi \in (0, \infty), \tilde \beta \in \mathbb{R}^d} \ \frac{1}{2n} \sum_{i = 1}^{n} \log f \left( y_i ; g^{-1} \left( \alpha + \sum_{j =1}^d \tilde \beta_j  k(x_j, x_i) \right), \phi \right) - \lambda \tilde \beta^T K_{d,d}^T \ \tilde \beta
\end{equation}
```

## Question 5

For this question we first need to obtain the penalty term as $\omega ^T \omega$ for some vector $\omega$. We can first obtain the eigen-decomposition of $K_{d,d}^T$:
$$
K_{d,d}^T = V \Lambda V^T
$$
Then writing $\Lambda = \Lambda^{\frac{1}{2}} \Lambda^{\frac{1}{2}} = \Lambda^{\frac{1}{2}} \left(\Lambda^{\frac{1}{2}}\right)^T$, we get
$$
\tilde \beta^T K_{d,d} \ \tilde \beta = \tilde \beta^T V \Lambda^{\frac{1}{2}} \left(\Lambda^{\frac{1}{2}}\right)^T V^T \tilde \beta = \omega^T \omega
$$

where $\omega^T = \tilde \beta^T V \Lambda^{\frac{1}{2}}$, or equivalently $\tilde \beta^T = \omega^T \Lambda^{- \frac{1}{2}} V^T$.

Then our objective function becomes 

```{=tex}
\begin{equation}
(\hat{\alpha_\lambda}, \hat{\phi_\lambda}, \hat \omega_\lambda ) \in \argmax{\alpha \in \mathbb{R}, \phi \in (0, \infty), \omega\in \mathbb{R}^d} \ \frac{1}{2n} \sum_{i = 1}^{n} \log f \left( y_i ; g^{-1} \left( \alpha +  \omega^T \Lambda^{- \frac{1}{2}} V^T K_{d,n} \right), \phi \right) - \lambda \omega^T \omega
\end{equation}
```

so we can use $X' = \Lambda^{- \frac{1}{2}} V^T K_{n,d}$, where $K$ is the Gram matrix of the original data, as the input for `glmnet()` and this will give the estimated $\omega$.

# Part 2

This section implements the theory from Part 1 on the `wesdr` data from the `gss` package, which has the binary response variable `ret` and explanatory variables `dur`, `gly` and `bmi`

```{r}
library(dplyr)
library(kableExtra)
library(gss)
data("wesdr")
head(wesdr)
```


## Question 6

We can now use `glmnet` to write a function implementing Question 4 and 5:

```{r}
library(kernlab)
estimate <- function(X, y, lambda, kernel, m){
  d <- m - 2
  K <- kernelMatrix(kernel = kernel, x = X)
  K_d <- K[1:d, 1:d]
  
  K_eigen <- eigen(K_d)
  V <- K_eigen$vectors
  Lambda <- (K_eigen$values)
  
  K_nd <- K[1:d,]
  X_new <- diag(1/sqrt(Lambda)) %*% t(V) %*% K_nd
  

  fit <- glmnet(t(X_new), y, family = "binomial", alpha = 0, lambda = lambda)
  
  alpha <- fit$a0
  omega <- fit$beta
  
  beta <- V %*% diag(1/sqrt(Lambda)) %*% (omega)
  
  return(c(numeric(alpha), as.vector(beta)))
}

```


```{r}
X <- as.matrix(wesdr[,1:3])
y <- wesdr[,4]



estimate(X, y, lambda = 0.3, kernel = rbfdot(sigma = 1), m = 300)
```




