---
title: "Parallel Rcpp"
author: "Rachel Wood"
date: "2023-04-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/home/ac18826/Documents/COMPASS/First_Year/TB2/Statistical_Computing/Parallel_Rcpp/")
```


```{r, engine='Rcpp'}
#include <unistd.h>
#include <Rcpp.h>

// [[Rcpp::export(wait_a_second)]]
bool wait_a_second(int sec)
{
 for(int ii = 0; ii < sec; ii++)
 { 
  sleep(1);
 }
 return 1;
}

```

```{r}
system.time( wait_a_second(2) )[3]
```

## Sleeping in Parallel
```{r, engine='Rcpp', verbose=TRUE}

#include <unistd.h>
#include <Rcpp.h>

// [[Rcpp::plugins(openmp)]]

// [[Rcpp::export(wait_a_second_omp)]]
bool wait_a_second_omp(int sec, int ncores)
{

 #if defined(_OPENMP)
  #pragma omp parallel num_threads(ncores)
  #pragma omp for
 #endif
 for(int ii = 0; ii < sec; ii++)
 { 
  sleep(1);
 }
 
 return 1;

 }
```


```{r}
system.time(wait_a_second_omp(4, 1))[3]
```

# Non-Linear Speed ups

```{r, engine='Rcpp'}
#include <Rcpp.h>
#include <omp.h>
using namespace Rcpp;

// [[Rcpp::export(allFiniteSeq)]]
bool allFiniteCpp(NumericVector x)
{

 size_t n = x.size();
 double out = 0;
 
 for(size_t ii = 0; ii < n; ii++)
 {
  out += x[ii];
 }
 
 return R_FINITE(out);

}

// [[Rcpp::plugins(openmp)]]

// [[Rcpp::export(allFiniteOMP)]]
bool allFiniteOMP(NumericVector x, int ncores)
{
 size_t n = x.size();
 double out = 0;
 NumericVector z(ncores);
 
 #pragma omp parallel num_threads(ncores)
 {
  int kk = omp_get_thread_num();
  
  #pragma omp for schedule(static)
  for(size_t ii = 0; ii < n; ii++)
  {
   z[kk] += x[ii];
  }
 }
 
 out = sum(z);
 
 return R_FINITE(out);

 }
```

```{r}
x <- rnorm(1e7)

library(microbenchmark)
options(microbenchmark.unit="relative")
microbenchmark(all(is.finite(x)), #Basic R
               allFiniteSeq(x), #Rcpp sequential
               allFiniteOMP(x, 1), #Rcpp and OpenMP ...
               allFiniteOMP(x, 4),
               allFiniteOMP(x, 8), 
               allFiniteOMP(x, 16))
```
Note: R API or Rcpp not thread-safe (opening a parallel section)

# Random Number Generation

The code below will produce invalid results and might crash your R session:
```{r, engine='Rcpp', eval = FALSE}
#pragma omp parallel
{
x = R::rbeta(a, b);
}
```

We adopt a thread-safe RNG
```{r, engine='Rcpp'}
#include <Rcpp.h>
#include <sitmo.h>

// [[Rcpp::depends(sitmo)]]

// [[Rcpp::export(sumunif_sitmo)]]
Rcpp::NumericVector sumunif_sitmo(unsigned int n,
                                  unsigned int nstep,
                                  double seed) {
  Rcpp::NumericVector out(n);

  uint32_t coreseed = static_cast<uint32_t>(seed);
  sitmo::prng eng(coreseed);
  
  double mx = sitmo::prng::max();
  double tmp = 0;

   for(unsigned int ii = 0; ii < n; ++ii) {
     tmp = 0.0;
     for(unsigned int kk = 0; kk < nstep; ++kk){
      tmp += eng() / mx;
     }
     out[ii] = tmp;
   }
    
  return out;
}
```
Equivalent to 

```{r, eval=FALSE}
rowSums(matrix(runif(n*nstep), n, nstep))
```


```{r}
n <- 1e3
nstep <- 1000

par(mfrow = c(1, 2))
hist(rowSums(matrix(runif(n*nstep), n, nstep)), main = "R", xlab = "x")
hist(sumunif_sitmo(n, nstep = nstep, seed = 1), main = "Sitmo", xlab = "x")
```

## Parallel Version

```{r, engine = 'Rcpp'}
#include <Rcpp.h>
#include <sitmo.h>

#ifdef _OPENMP
#include <omp.h>
#endif

// [[Rcpp::depends(sitmo)]]
// [[Rcpp::plugins(openmp)]]

// [[Rcpp::export(sumunif_sitmo_omp)]]
Rcpp::NumericVector sumunif_sitmo_omp(unsigned int n,
                                      unsigned int nstep,
                                      Rcpp::NumericVector seeds) {
  Rcpp::NumericVector out(n);
  
  unsigned int ncores = seeds.size();
  
  #ifdef _OPENMP
  #pragma omp parallel num_threads(ncores)
  {
  #endif
  
   uint32_t coreseed = static_cast<uint32_t>(seeds[0]);
   
   #ifdef _OPENMP
    coreseed = static_cast<uint32_t>(seeds[omp_get_thread_num()]);
   #endif
   
   sitmo::prng eng(coreseed);
   
   double mx = sitmo::prng::max();
   double tmp = 0;
  
   #ifdef _OPENMP
    #pragma omp for 
   #endif
   for(unsigned int ii = 0; ii < n; ++ii) {
     tmp = 0.0;
     for(unsigned int kk = 0; kk < nstep; ++kk){
      tmp += eng() / mx;
     }
     out[ii] = tmp;
   }
    
  #ifdef _OPENMP
  }
  #endif
  
  return out;
}
```
Compare results:
```{r}
microbenchmark(R = rowSums(matrix(runif(n*nstep), n, nstep)),
               sitmo = sumunif_sitmo(n, nstep = nstep, seed = 1),
               sitmo_omp1 = sumunif_sitmo_omp(n, nstep = nstep, seeds = 1),
               sitmo_omp4 = sumunif_sitmo_omp(n, nstep = nstep, seeds = 1:4), 
               sitmo_omp16 = sumunif_sitmo_omp(n, nstep = nstep, seeds = 1:16), 
               times = 100
               )
```
# Thread Safe Accessors

Consider the error function
```{r}
x <- rnorm(1e5)
2 * pnorm(x * sqrt(2)) - 1
```

The same in Rcpp is 
```{r, engine='Rcpp'}
#include <boost/math/special_functions/erf.hpp>
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::depends(BH)]]

// [[Rcpp::export(erf)]]
NumericVector erf(NumericVector x)
{

 size_t n = x.size();
 NumericVector out(n);
 
 for(size_t ii = 0; ii < n; ii++)
 {
  out[ii] = boost::math::erf(x[ii]);
 }
 
 return out;

 }
```

check it works:

```{r}
x <- rnorm(1e6)
max(abs( (2 * pnorm(x * sqrt(2)) - 1) - erf(x)))

```

```{r}
options(microbenchmark.unit="relative")
microbenchmark(R = 2 * pnorm(x * sqrt(2)) - 1, 
               erf(x), 
               times = 100)
```


Parallelising the computation:
```{r, engine='Rcpp'}
#include <boost/math/special_functions/erf.hpp>
#include <Rcpp.h>
#include <RcppParallel.h>
using namespace Rcpp;

// [[Rcpp::depends(BH)]]
// [[Rcpp::depends(RcppParallel)]]
// [[Rcpp::plugins(openmp)]]

// [[Rcpp::export(erfOmp)]]
NumericVector erfOmp(NumericVector x, int ncores)
{

 size_t n = x.size();
 NumericVector out(n);
 RcppParallel::RVector<double> wo(out);
 RcppParallel::RVector<double> wx(x);
 
 #if defined(_OPENMP)
  #pragma omp parallel for num_threads(ncores)
 #endif
 for(size_t ii = 0; ii < n; ii++)
 {
  wo[ii] = boost::math::erf(wx[ii]);
 }
 
 return out;

 }
```

```{r}
microbenchmark(R = 2 * pnorm(x * sqrt(2)) - 1, 
               erfOmp(x, 1), 
               erfOmp(x, 4),
               erfOmp(x, 8),
               times = 100)

```

