---
title: "Parallel Rcpp Portfolio"
author: "Rachel Wood"
date: "2023-04-20"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For this portfolio we consider the `electBook` dataset on Irish smart meters:
```{r}
library(electBook)

data("Irish")
summary(Irish) 
```

We concatenate the household demand into a vector - as there are a huge number of observations, we only plot a sub sample of the data:
```{r}
y <- do.call("c", Irish$indCons)
y <- y - mean(y)
```

```{r}
ncust <- ncol(Irish$indCons) 

x <- rep(Irish$extra$tod, ncust)

n <- length(x)
ss <- sample(1:n, 1e4)
plot(x[ss], y[ss], col = "grey")
```
## Modelling Demand
We consider modelling the demand as a function of the time of day. Our first model is a simple linear regression $\mathbb{E} (y|x) = \beta x$, which we fit using least squares:


```{r}
reg1D <- function(y, x){
  
  b <- t(x) %*% y / (t(x) %*% x)  
  
  return(as.vector(b))
  
}
```

We compare this to the built in `lm` function:
```{r}
system.time( lm(y ~ -1 + x)$coeff )[3]

system.time( reg1D(y, x) )[3]
```

We can see our function has a much faster implementation, however we can still attempt to improve this by parallelising the code in Rcpp using OpenMP
```{r, engine='Rcpp'}

#include <Rcpp.h>
#include <omp.h>
using namespace Rcpp;

// [[Rcpp::plugins(openmp)]]
// [[Rcpp::export]]
double reg1DParallel(NumericVector y, NumericVector x) {
  int n = y.size();
  double numerator = 0.0;
  double denominator = 0.0;

  #pragma omp parallel for reduction(+:numerator, denominator)
  for (int i = 0; i < n; i++) {
    numerator += x[i] * y[i];
    denominator += x[i] * x[i];
  }

  return numerator / denominator;
}

```

We first check the results are the same:


```{r}
all.equal(reg1D(y,x), reg1DParallel(y,x))
```
and now compare computational time:

```{r}
system.time( reg1D(y, x) )[3]
system.time( reg1DParallel(y, x) )[3]
```
We can see this has improved by a factor of 10.





