---
title: "Parallel Rcpp Portfolio"
author: "Rachel Wood"
date: "2023-04-20"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For this portfolio we consider the `electBook` dataset on Irish smart meters:
```{r}
library(electBook)
library(kableExtra)
data("Irish")
summary(Irish) %>%
  kable()
```

We concatenate the household demand into a vector - as there are a huge number of observations, we only plot a sub sample of the data:
```{r}
y <- do.call("c", Irish$indCons)
y <- y - mean(y)
```

```{r}
ncust <- ncol(Irish$indCons) 

x <- rep(Irish$extra$tod, ncust)

n <- length(x)
ss <- sample(1:n, 1e4)
plot(x[ss], y[ss], col = "grey")
```
## Modelling Demand
We consider modelling the demand as a function of the time of day. Our first model is a simple linear regression $ \mathbb{E} (y|x) = \beta x$, which we fit using least squares:


```{r}
reg1D <- function(y, x){
  
  b <- t(x) %*% y / (t(x) %*% x)  
  
  return(b)
  
}
```

We compare this to the built in `lm` function:
```{r}
system.time( lm(y ~ -1 + x)$coeff )[3]

system.time( reg1D(y, x) )[3]
```

We can see our function has a much faster implementation, however we can still attempt to improve this by parallelising the code in Rcpp 
```{r, engine='Rcpp'}
#include <Rcpp.h>
#include <RcppParallel.h>
using namespace Rcpp;

// [[Rcpp::depends(RcppParallel)]]
// [[Rcpp::plugins(openmp)]]

// [[Rcpp::export]]
int reg1D_RcppParallel(NumericVector y, NumericVector x){
  
}
```

