---
title: "Comparing CLARITY, Omnibus Embedding and UASE"
author: "Rachel Wood"
date: "2023-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
set.seed(124)
library(ggplot2)
theme_set(theme_bw())
```

This document's goals are three-fold:

- Looking at how well different methods perform in identifying changes in the structure of data.
- Finding the way UASE embeddings change with scaling and mixing changes to the underlying generating tree structure.
- Identify a performance metric with which we can compare these different methods

# Data generation

We use the `ClaritySim` package to create the `original' data generated according to some tree structure, which we will then manipulate to test the different methods:

```{r}
library(ClaritySim)
n=100 
k=10
original <- simulateCoalescent(n,k,
                         sigma0=0.0001,
                         Amodel="uniform",
                         alpha=0,
                         minedge=0.1) # Simlulate 100 objects in a 10 dimensional latent space
plot(original$tree)
```

We can now manipulate the data in a series of ways.



## Tree Scaling

```{r}
scaled <- transformCoalescent(original)
plot(scaled$tree)
```

## Mixture

```{r}
mixed <- mixCoalescent(original, fraction = 1)
mixture <- mixed$edges

library(ape)
library(glottoTrees)

plot(mixed$tree)
edges(mixture[2], mixture[1], arrows = 1)

```

## Swapping node - REVISIT

```{r}
tree <- original$tree

newtree <- tree
newtree$tip.label <-c("t2", "t1", "t3", "t4", "t5", "t6", "t7", "t8", "t9", "t10")

swapped <- simulateCoalescent(n,k,
                         sigma0=0.0001,
                         Amodel="uniform",
                         alpha=0,
                         minedge=0.1,
                         tree = newtree) 
plot(swapped$tree)
```


# UASE

We first include our R functions for UASE embedding
```{r}
library(ggplot2)
source("UASE.R")
library(adephylo)
scaled_UASE <- UASE(original, scaled)

plot_UASE(scaled_UASE)

distRoot(original$tree)
```





