---
title: "Comparing CLARITY, Omnibus Embedding and UASE"
author: "Rachel Wood"
date: "2023-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
set.seed(124)
library(ggplot2)
theme_set(theme_bw())
```

This document's goals are three-fold:

- Looking at how well different methods perform in identifying changes in the structure of data.
- Finding the way UASE embeddings change with scaling and mixing changes to the underlying generating tree structure.
- Identify a performance metric with which we can compare these different methods

# Data generation

We use the `ClaritySim` package to create the `original' data generated according to some tree structure, which we will then manipulate to test the different methods:

```{r}
library(ClaritySim)
n=100 
k=10
original <- simulateCoalescent(n,k,
                         sigma0=0.1,
                         Amodel="uniform",
                         alpha=0,
                         minedge=0.1) # Simlulate 100 objects in a 10 dimensional latent space
plot(original$tree)
```

We can now manipulate the data in a series of ways.

## Adding an Anomaly:

```{r}
A <- original$A
A[95:100,] <- A[85:90,]

anomaly <- original
anomaly$A <- A
anomaly <- simData(anomaly)
source("UASE.R")

anomaly_UASE <-UASE(original, anomaly, d = 8)
plot_UASE(anomaly_UASE)

distances <- apply(as.matrix(1:8, nrow = 1), MARGIN = 1, FUN = function(ii){
  distance_moved(anomaly_UASE, d = ii)
})
distances <- as.data.frame(distances) 

distances <- distances %>% tibble(.name_repair = "minimal")
colnames(distances) <- paste("dim", as.character(1:8))
heatmap(distances, Rowv = NA, Colv = NA)
```


## Tree Scaling
We now look at how tree scaling affects the detection of anomalies, we will set the minimum and maximum scaling coefficient to $\frac{1}{r}$ and $r$ respectively for different values of $r$:
```{r}
r <- as.vector(1:10)
scaled <- vector(mode = "list", length(r))
names(scaled) <- r

for (i in 1:length(r)){
  scaled[[i]] <- transformCoalescent(anomaly, multmin = 1/r[i], multmax = r)
}

scaled_UASE <- lapply(scaled, function(sim){UASE(original, sim)})
```

## Mixture

```{r}
set.seed(2)
mixed <- mixCoalescent(original, fraction = 1, transform = TRUE)
mixture <- mixed$edges

library(ape)
plot(mixed$tree)
edges(mixture[1], mixture[2], arrows = 1)

```

```{r}
test=original
test$Y=test$Y
mixed_UASE <-UASE(test, mixed, d = 12)
plot_UASE(mixed_UASE)

```

```{r}
library(hrbrthemes)
distances <- apply(as.matrix(1:12, nrow = 1), MARGIN = 1, FUN = function(ii){
  distance_moved(mixed_UASE, d = ii)
})

distance_plot <- expand.grid( Components = as.factor(1:12), Observations = 1:nrow(distances))
distance_plot$distances <- c(t(distances))
distance_plot$groups <- trunc( (distance_plot$Observations-1)/10) +1


distance_plot <- as_tibble(distance_plot)
ggplot(distance_plot, aes(x = Components))+
  geom_tile(aes(y = Observations, fill = distances)) +
  theme_ipsum()+
  scale_fill_distiller(direction = 1)
```

# UASE

We first include our R functions for UASE embedding
```{r, eval = FALSE}
library(ggplot2)
source("UASE.R")
scaled_UASE <- UASE(original, scaled)

plot_UASE(scaled_UASE)

#original_root_dist <- distRoot(original$tree)
#scaled_root_dist <- distRoot(scaled$tree)


```





