---
title: "Comparing CLARITY, Omnibus Embedding and UASE"
author: "Rachel Wood"
date: "2023-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
set.seed(124)
library(ggplot2)
theme_set(theme_bw())
```

This document's goals are three-fold:

- Looking at how well different methods perform in identifying changes in the structure of data.
- Finding the way UASE embeddings change with scaling and mixing changes to the underlying generating tree structure.
- Identify a performance metric with which we can compare these different methods

# Data generation

We use the `ClaritySim` package to create the `original' data generated according to some tree structure, which we will then manipulate to test the different methods:

```{r}
library(ClaritySim)
n=100 
k=10
original <- simulateCoalescent(n,k,
                         sigma0=0.1,
                         Amodel="uniform",
                         alpha=0,
                         minedge=0.1) # Simlulate 100 objects in a 10 dimensional latent space
plot(original$tree)
```

We can now manipulate the data in a series of ways.

## Adding an Anomaly:

```{r}
A <- original$A
A[95:100,] <- A[85:90,]

anomaly <- original
anomaly$A <- A
anomaly <- simData(anomaly)
source("UASE.R")

anomaly_UASE <-UASE(original, anomaly)
plot_UASE(anomaly_UASE)
```


## Tree Scaling
We now look at how tree scaling affects the detection of anomalies, we will set the minimum and maximum scaling coefficient to $\frac{1}{r}$ and $r$ respectively for different values of $r$:
```{r}
r <- as.vector(1:10)
scaled <- vector(mode = "list", length(r))
names(scaled) <- r

for (i in 1:length(r)){
  scaled[[i]] <- transformCoalescent(anomaly, multmin = 1/r[i], multmax = r)
}

scaled_UASE <- lapply(scaled, function(sim){UASE(original, sim)})
```

## Mixture

```{r}
mixed <- mixCoalescent(original, fraction = 1)
mixture <- mixed$edges

library(ape)
library(glottoTrees)

plot(mixed$tree)
edges(mixture[2], mixture[1], arrows = 1)

```

## Swapping node - REVISIT

```{r, eval = FALSE}
tree <- original$tree

newtree <- tree
newtree$tip.label <-c("t2", "t1", "t3", "t4", "t5", "t6", "t7", "t8", "t9", "t10")

swapped <- simulateCoalescent(n,k,
                         sigma0=0.0001,
                         Amodel="uniform",
                         alpha=0,
                         minedge=0.1,
                         tree = newtree) 
plot(swapped$tree)
```


# UASE

We first include our R functions for UASE embedding
```{r, eval = FALSE}
library(ggplot2)
source("UASE.R")
scaled_UASE <- UASE(original, scaled)

plot_UASE(scaled_UASE)

#original_root_dist <- distRoot(original$tree)
#scaled_root_dist <- distRoot(scaled$tree)


```





